Constructing a universe for the setoid model

### 2020-06-30 paper.tex
Changed institution.
More paper writing

\author{
  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Simon Boulier
\and
  Ambrus Kaposi \inst{2}
\and
  Christian Sattler\inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Filippo Sestini\inst{1}
}

\institute{
  School of Computer Science, University of Nottingham, UK\\
  \email{\{psztxa,pszcs1,psxfs5\}@nottingham.ac.uk}
  \and
  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary\\
  \email{akaposi@inf.elte.hu}
}

@@

A central aspect of Type Theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand ``propositional'' equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics in type theory, one should aim for
a notion of definitional equality that is decidable and therefore
machine-checkable, and a notion of propositional equality that is as as
extensional possible.

Two extensionality principles seem particularly desirable, in that they arguably
constitute the bare minimum for type theory to be comparable to set theory as a
foundational system for set-level mathematics, in terms of power and ergonomics.
One is function extensionality (funext), according to which functions are equal
if point-wise equal. Another is propositional extensionality (propext), equating
all propositions that are logically equivalent.

Extensional Type Theory (ETT) does support extensional reasoning to some degree,
and thus would seem a convenient setting for formalization of
mathematics. However, ETT's equality reflection makes type-checking ETT terms
computationally unfeasible, if not impossible.

On the other hand, Intensional Type Theory (ITT) has nice computational
properties that can make it more suitable for computer implementation, but as
normally defined [cite Martin-L\"of paper?] it lacks extensionality.
%
It is known from model constructions that extensional principles like funext are
consistent with ITT.
%
Moreover, ITT extended with uniqueness of identity proofs (UIP) and funext is
known to be as powerful as ETT [cite hofmann]. These principles can be added as
axioms, however destroying some computational properties of ITT such as
canonicity.

What we would like instead is a formulation of ITT that supports the
aforementioned extensionality principles, while retaining convenient
computational properties like canonicity and decidability of type checking.
%
Unfortunately, one can show that whenever identity types are defined according
to Martin-Loef's inductive definition, there is no hope to have canonicity and
funext coexist.
%
The first step towards a solution is to give up the idea of propositional
equality as single type family inductively generated by reflexivity. Instead,
equality should be \emph{specific} to each type former in the type theory, or in
other words, every type former should be introduced alongside an explanation of
what counts as equality for its elements. In this setting, function
extensionality can simply be introduced as the chosen notion of equality on
function types.

This conception of types together with their notion of equality goes back to the
notion of \emph{setoids} or \emph{Bishop sets}. A setoid is simply a set
equipped with a propositional equivalence relation. Setoids provide a quite
natural and useful semantic domain in which to interpret type theory. The first
setoid model was constructed to show consistency of funext for ITT [cite]. This
consistency result is implied by the observation that UIP and funext hold in the
setoid model. Moreover, it was shown by Altenkirch [cite] that is the model
construction is carried out in a metatheory with a universe of strict
(definitionally proof-irrelevant) propositions, it is possible to define a
univalent universe of propositions, therefore satisfying propositional
extensionality. The setoid model thus satisfies all the extensionality
principles that we would like to extend ITT with. The question is whether there
exists a version of ITT that supports setoid reasoning, and hence the forms of
extensionality enables by it.

This question was revisited and answered in Altenkirch et al. [cite]. In this
paper, the authors define SeTT an extension of intensional MLTT + Prop with
constructs for setoid reasoning, where funext and propext hold by
definition. Moreover, they justify SeTT via a syntactic translation, after
observing that this is made possible by the strictness of the model (every
equation holds definitionally). This is in contrast with other type theories
based on the setoid model, like OTT [cite] and XTT [cite], which instead rely on
ETT as metatheory for their justification. A major contribution of SeTT is
therefore to provide a way to bootstrap extensionality, by translating it into a
small core intensional theory by means that do not require extensionality to
begin with.

The current formulation of SeTT is already a rich theory. However, its
introspection capabilities are currently lacking, as they are limited to the
univalent universe of propositions. We would like to internalise the notion of
type in SeTT, thus extending the theory with a universe of setoids.
%
This brings up several questions, one of which has to do with the notion of
equality with which the universe should come equipped. The universe of setoids
is itself a setoid (as any type is), so it certainly cannot be a univalent
universe, since a univalent setoid would be a groupoid rather than a setoid.
%
Another question asks how one can justify such universe in the setoid model. In
particular what principles, are needed from the metatheory in order to define
the universe.

@@

This paper documents our work towards the construction of a universe of setoids
inside the setoid model, and tries to answer these and other questions related
to the design and implementation of this construction. Our main contribution is
the definition of the universe in the setoid model; this is initially given as
an inductive-inductive-recursive definition, which is then translated firstly to
an inductive-inductive definition, and secondly to a purely inductive type. As a
consequence, we show that we only need to assume indexed W-types in the
metatheory (along with some obligatory basic tools like $\Sigma$ and $\Pi$
types) to justify the universe.

@@

We begin by describing the metatheory that we will use throughout the paper, in
Section~\ref{sec:metatheory}. In Section~\ref{sec:setoid-model}, after briefly
recalling the abstract notion of models of Type Theory known as categories with
familes (CwF), we outline Altenkirch's setoid model of Type Theory in the
variant used in \cite{mpc19}. We then show how the model has been turned into
Setoid Type Theory in Section~\ref{sett}.

In Section~\ref{setoid-universe} we provide the actual construction of the
setoid universe in the setoid model. In Section~\ref{ir-universes} we first
recall the well known inductive-recursive definition of universes in Type
Theory, and the way these can be turned into a construction that does not
require induction-recursion.
%
We then provide, in Section~\ref{ir-setoid-universe}, a first complete
definition of the setoid universe using a quite complex form of
induction-recursion. This is followed by an alternative definition of the
universe in Section~\ref{ii-universe}, that does not rely on induction-recursion
but instead on infinitary induction-induction.
%
Further reductions of this inductive-inductive universe to even simpler forms,
in addition to other long term goals, are discussed in
Section~\ref{further-work}.

@@

TODO

@@

Note that this
represents a strong elimination principle, as it allows to eliminate a strict
proposition into arbitrary types. Indeed, the reason why we must assume this
identity type as primitive is because, even though we could have defined
\textsf{Id} from $W_p$, \textsf{transp} would not be definable from $W_p$
elimination. \todo{what is the justification for this?}

@@

A universe of strict propositions has been recently added to the Agda proof
assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact, most of
the developments presented here have been formalized and proof-checked using
Agda extended with strict propositions.
%
For convenience, we also slightly deviate from $MLTT^{Prop}$ both in the paper
and in the formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
instead of their derived forms. We operate under the assumption that everything
can be equivalently carried out in $MLTT^{Prop}$, although we haven't fully
checked all the necessary details.

@@

%% The setoid interpretation gives rise to a model of a type theory with functional
%% and propositional extensionality. It is thus a way to explain extensionality in
%% a type-theoretic and computationally adequate way \cite{setoid-model}.

By \emph{setoid model} we intend a class of models of Type Theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional Type Theory by M. Hofmann in
\cite{hofmann-ext-int}, in order to give a semantics to extensionality
principles such as function and propositional extensionality. Here we consider
a similar model construction due to Altenkirch \cite{setoid-model}.
%
The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory which includes a strict universe of propositions.
%
The use of strict propositions to define the setoid equivalence relations makes
this model \emph{strict}. A strict model is one where every equation holds
definitionally. The setoid model thus defined validates function extensionality,
a univalent universe of propositions, and quotient types. Therefore, it provides
a way to bootstrap and ``explain'' extensionality, since the model construction
effectively gives an implementation of various extensionality principles it in
terms of a small, entirely intensional type theory.

@@

As any model of Type Theory, the setoid model can be framed categorically as a
category with families (CwF) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:

@@

This definition of types in the setoid model is different from the one in [cite
  LICS 99 paper], but it is equivalent to it [cite]. The main difference here is
in the use of heterogeneous equality in the definition of types.

@@

As defined, we can show that the setoid model validates the usual basic type
formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
of strict propositions satisfying propositional extensionality [cite LICS
  99]. Note that we don't need neither identity types nor inductive types
(W-types) in the metatheory to achieve this. However, W-types are needed when
constructing the setoid universe in the next sections.

@@

The setoid model presented in the previous section is peculiar in several
ways. One is that it can be presented in a very small type theoretic
metatheory. The other important aspect is that it is \emph{strict}, namely every
equation of the original syntax holds by definition in the semantics. One
advantage of strict models is that they can be turned into \emph{syntactic
  translations}, in which syntactic objects of the source theory are interpreted
as their counterparts in another \emph{target} theory. In the case of the setoid
model, this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.\todo{maybe expand this?}%
\footnote{%
Semantically, this translation corresponds to a model construction, in particular a functor from the category of models of the target theory to the category of models of what will be Setoid Type Theory.
Since the setoid translation is structural in the context component, we can work with models in the style of categories with families rather than contextual categories.
}

@@

Note that contractibility of singletons holds in SeTT by UIP and the computation
rules of equality on $\Sigma$ types. Hence, the J eliminator can also be
derived.

@@

In its current form, SeTT is already rich and interesting in its own right.
However, as pointed out in the introduction, it is also seriously limited by the
lack of a universe internalizing the notion of types beyond propositions, that
is, a universe of setoids. Our goal is to extend SeTT with such universe; since
SeTT is a direct syntactic reflection of the setoid model, this essentially
amounts to showing that a universe of setoids with the necessary structure and
equations can be constructed within the setoid model. This opens several
questions and possible design choices.

A first fundamental consideration has to with the very definition of the setoid
universe: as any type in the setoid model, this universe must be a setoid and
thus come equipped with an equivalence relation. However, unlike the already
existing universe of propositions, a universe of setoids cannot be univalent,
since this would force it to be a groupoid. Instead, we can define the relation
so that it reflects a simpler syntactic equality of codes.

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section TODO and shown originally in [cite LICS
  99] is that it can be carried out in a very small type theoretic metatheory,
thus providing a way to reduce extensionality to a small intensional core.
%
In extending the setoid model with a universe, we would like this property to
still hold, hence it is of primary concern to reduce to a minimum the extensions
required in the metatheory, so to keep it as small as possible.

A known and established method for constructing universes in Type Theory uses
induction-recursion (IR), a definition schema developed by Dybjer
[cite]. Inductive-recursive definitions can be found throughout the literature
on type theory, from the already mentioned type theoretic universes, including
the original formulation a la Tarski by Martin-L\"of [cite], to metamathematical
tools like computability predicates [cite].

Although universe constructions in Type Theory---including our own setoid
universe---are naturally presented as inductive-recursive definitions, they
may not necessarily require a metatheory that supports induction-recursion. In
fact, it is possible to reduce some instances of induction-recursion to plain
induction (more specifically, inductive families), including some universe
definitions. We recall this reduction in Section TODO.

%% However, we would prefer not to be forced to assume inductive-recursive types,
%% or IR in general, in the basic type theory which is the target of the setoid
%% translation. This is because the point of the translation is to explain a
%% complex system in terms of a very simple, well understood core, and we think
%% introducing IR would go against this goal.

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support elimination, thus exposing the inductive structure of
the codes. Universe elimination is known to be inconsistent with univalence. As
it turns our, univalence is not an issue in our case; nevertheless it is not
immediately clear if the elimination principle can be justified by the
semantics.

Another choice is whether to include codes for other universes, besides the
usual type formers. One could have $\Prop$ be a code in the setoid universe,
rather than a type. Moreover, the universes could form a hierarchy, where the
universe at level $i$ is an element of the universe of level $i+1$. Another
option is whether to make the hierarchy cumulative.

%% In the next section, we recall the inductive-recursive construction of universes
%% in type theory, as well as lay out the process by which such construction can be
%% turned to one that only relies on inductive families. In Section TODO, we give a
%% first inductive-recursive definition of the universe of setoids. After that we
%% will see if and how such IR definition can be turned into one that does not rely
%% on IR.

@@

%% Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
%% because... \todo{TODO add explanation}.

Induction-recursion is arguably the nicest and more natural way to define
internal universes in Type Theory, however it is not always required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$).

Following this, we seen that we can actually encode the universe above as an
inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
in the image of $\El$. In other words, $\ad{in-U}$ is a predicate that holds for
any type that would have been obtained via $\El$ in the usual
inductive-recursive definition. As $\El$ is indexed on the type of codes, the
definition of $\ad{in-U}$ quite expectedly reflecs the inductive structure of
codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible sets. This
also means that a metatheory with at least one universe is required for this
construction.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling Section TODO,
these are essentially record types made of several components. Since $\Uty$ is a
closed type, it requires the same data of a setoid; in particular, we need a type
of codes together with an equivalence relation reflecting equality of codes:
%
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
  %% & \El : \setoidU \to \Set \\
  %% & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}
%
in addition to proofs that these are indeed equivalence relations
%
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}

To define $\EL$, we need a family of setoids indexed over the universe. In other
words, a way to assign to each code in the universe a carrier set and an
equivalence relation.
%
\begin{align*}
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

@@

This gives a complete definition of a universe of setoids. However, as already
pointed out, this definition uses induction-recursion. Moreover, the kind of
recursion involved in this definition is particularly complex, since TODO. It is
not clear whether this definition can be reduced to simpler, better understood
notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
need to extend the metatheory with simple forms of induction-recursion goes
against our objective to keep the metatheory as small and essential as possible.

>>>>

\author{
  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Simon Boulier
\and
  Ambrus Kaposi \inst{2}
\and
  Christian Sattler\inst{3}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
\and
  Filippo Sestini\inst{1}
}

\institute{
  School of Computer Science, University of Nottingham, UK\\
  \email{\{psztxa,pszcs1,psxfs5\}@nottingham.ac.uk}
  \and
  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary\\
  \email{akaposi@inf.elte.hu}
  \and
  Chalmers University of Technology, Gothenburg, Sweden
  \email{sattler@chalmers.se}
}

@@

Intuitionistic type theory is a formal system designed by Per Martin-L\"of to be
a full-fledged foundation in which to develop constructive mathematics
\cite{mltt75, libretto}.
%
A central aspect of type theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand ``propositional'' equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics, one should aim for a notion of
propositional equality that is as \emph{extensional} as possible.

Two extensionality principles seem particularly desirable, since they arguably
constitute the bare minimum for type theory to be comparable to set theory as a
foundational system for set-level mathematics, in terms of power and ergonomics.
One is function extensionality (or \emph{funext}), according to which functions
are equal if point-wise equal. Another is propositional extensionality
(or \emph{propext}), that equates all propositions that are logically equivalent.

Type theory with equality reflection, also known as \emph{extensional type
  theory} (ETT) does support extensional reasoning to some degree, but
unfortunately equality reflection makes the problem of type-checking ETT terms
computationally unfeasible, if not impossible.

On the other hand, \emph{intensional type theory} (ITT) has nice computational
properties like decidable type checking that can make it more suitable for
computer implementation, but as normally defined (for example, in \cite{mltt75})
it severely lacks extensionality.
%
It is known from model constructions that extensional principles like funext are
consistent with ITT.
%
Moreover, ITT extended with the principle of \emph{uniqueness of identity
  proofs} (UIP) and funext is known to be as powerful as ETT
\cite{conservativity}. We could recover the expressive power of ETT by adding
these principles to ITT as axioms, however destroying some computational
properties like canonicity.

What we would like instead is a formulation of ITT that supports extensionality,
while retaining its convenient computational behaviour.
%
Unfortunately, one can show that canonicity and function extensionality cannot
coexist if propositional equality is given by Martin-L\"of's inductively-defined
identity types.
%
The first step towards a solution is to give up the idea of propositional
equality as a single inductive definition given homogeneously for an arbitrary
type. Instead, equality should be \emph{specific} to each type former in the
type theory, or in other words, every type former should be introduced alongside
an explanation of what counts as equality for its elements.
%% In this setting, function extensionality can simply be introduced as the
%% chosen notion of equality on function types.

This idea of pairing types together with their own equality relation goes back
to the notion of \emph{setoid} or \emph{Bishop set}. Setoids provide a quite
natural and useful semantic domain in which to interpret type theory. The first
setoid model was constructed to show consistency of funext for ITT
\cite{hofmann}. This consistency result is implied by the observation that UIP
and funext hold in the setoid model. Moreover, it was shown by Altenkirch
\cite{setoid99} that is the model construction is carried out in a type
theoretic metatheory with a universe of strict (definitionally proof-irrelevant)
propositions, it is possible to define a univalent universe of propositions
satisfying propositional extensionality. The setoid model thus satisfies all the
extensionality principles that we would like to have in type theory. The
question is whether there exists a version of intensional type theory that
supports setoid reasoning, and hence the forms of extensionality enabled by it.

This question was revisited and answered in Altenkirch et al. \cite{mpc19}. In
this paper, the authors define Setoid Type Theory (SeTT), an extension of
intensional Martin-L\"of type theory with constructs for setoid reasoning, where
funext and propext hold by definition. SeTT is based on the \emph{strict} setoid
model of Altenkirch\footnote{a \emph{strict} model is one where every equation
  holds definitionally}, which makes it possible to show consistency via a
syntactic translation. This is in contrast with other type theories based on the
setoid model, like Observational Type Theory \cite{alti:ott-conf, alti:ott-conf}
and XTT \cite{xtt}, which instead rely on ETT for their justification. A major
property of SeTT is thus to illustrate how to bootstrap extensionality, by
translation into a small intensional core.

SeTT as defined in \cite{mpc19} is already a rich theory, but its introspection
capabilities are currently lacking, as its universes are limited to
propositions. We would like to internalise the notion of type in SeTT, thus
extending the theory with a universe of setoids.
%
This goal brings up several questions, one of which has to do with the notion of
equality with which the universe should come equipped: the universe of setoids
is itself a setoid (as any type is) so it certainly cannot be univalent, since
setoids lack the necessary structure.
%
Another issue is the way such universe can be justified by the setoid model, and
in particular what principles are needed in the metatheory to do so.

@@

This paper documents our work towards the construction of a universe of setoids
inside the setoid model, and tries to answer these and other questions related
to the design and implementation of this construction. Our main contribution is
the construction of the universe in the model; this is given in steps, first as
an inductive-recursive definition, which is then translated to an
inductive-inductive definition, and subsequently to an inductive type. As a
consequence, we show that we only need to assume indexed W-types and
proof-irrelevant identity types in the metatheory (along with some obligatory
basic tools like $\Sigma$ and $\Pi$ types) to construct the universe.

The universe constructions presented in this paper are, to our knowledge, the
first examples of two kinds of data type reductions in an intensional
metatheory: the first involving an inductive-recursive type which including
strict propositions, and the second involving an infinitary inductive-inductive
type.

@@

We begin by describing the metatheory that we will use throughout the paper, in
Section~\ref{sec:metatheory}. In Section~\ref{sec:setoid-model}, after briefly
recalling \emph{categories with families} as an abstract notion of models of
type theory, we outline Altenkirch's setoid model as given in \cite{mpc19}. We
then briefly discuss the rules of Setoid Type Theory in Section~\ref{sett}.

In Section~\ref{setoid-universe} we discuss the setoid model and various design
choices related to it. We then recall inductive-recursive universes, and the way
they can be equivalently defined as a plain inductive definition, in
Section~\ref{ir-universes}.
%
We then provide, in Section~\ref{ir-setoid-universe}, a first complete
definition of the setoid universe using a form of induction-recursion. This is
followed by an alternative definition of the universe in
Section~\ref{ii-universe}, that does not rely on induction-recursion but instead
on infinitary induction-induction. This inductive-inductive encoding of the
universe is obtained from the inductive-recursive one, following the method of
Section~\ref{ir-universes}.
%
We end the series of universe constructions with
Section~\ref{inductive-universe}, where we outline a purely inductive definition
of the setoid universe, obtained from the inductive-inductive one.

We conclude the paper with a discussion of further work in
Section~\ref{further-work}.

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf, alti:ott-conf} is a syntax for the setoid
model differing from SeTT for the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.

The principle of propositional extensionality in the setoid model is an instance
of Voevodsky's univalence axiom \cite{hottbook}. The cubical set model is a
constructive model justifying this axiom \cite{cubical-sets}. A type theory
extracted from this model is Cubical Type Theory \cite{cubical}. The
relationship between the cubical set model and cubical type theory is similar to
that between the setoid model and SeTT.
%
Compared to cubical type theories, SeTT has the advantage that the equality type
satisfies more definitional equalities. For instance, whereas in cubical type
theory equality of functions is isomorphic to pair-wise equality, in SeTT the
isomorphism is replaced by a definitional equality. SeTT is also a syntactically
straightforward extension of Martin-L\"of Type Theory, that does not require
exotic objects like the internal pretype. In turn, the obvious advantage of
cubical type theory is that it is not limited to setoids.

An exceptional aspect of the metatheory used in this paper is the presence of a
proof-irrelevant identity type with a strong transport rule allowing to
eliminate into arbitrary types.
%
In \cite{abel:nbe09}, Abel given a proof of normalization by evaluation for the
Logical Framework extended with a similar proof-irrelevant equality type.
%
Abel and Coquand show in \cite{abel2019failure} that the combination of
impredicativity with a strong transport rule results in terms that fail to
normalize.

@@

Note that \textsf{transp} provides a strong elimination principle allowing to
eliminate a strict proposition (the identity type) into arbitrary types. Indeed,
the reason why we must assume this identity type as primitive is because, even
though we could have defined \textsf{Id} from $W_p$, \textsf{transp} would not
be definable from $W_p$ elimination. \todo{what is the justification for this?}

@@

A universe of strict propositions has been recently added to the Agda proof
assistant \cite{agda-prop}, thus making $MLTT^{Prop}$ a subset of Agda. In fact,
most of the developments presented here have been formalized and proof-checked
using Agda extended with strict propositions.
%
For convenience, we slightly deviate from $MLTT^{Prop}$ both in the paper and in
the formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $MLTT^{Prop}$, although we
haven't fully checked all the necessary details.

@@

By \emph{setoid model} we intend a class of models of Type Theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional Type Theory by M. Hofmann in
\cite{hofmann}, in order to give a semantics to extensionality
principles such as function and propositional extensionality. Here we consider
a similar model construction due to Altenkirch \cite{setoid99}.
%
The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory which includes a strict universe of propositions.
%
The use of strict propositions to define the setoid equivalence relations makes
this model \emph{strict}. A strict model is one where every equation holds
definitionally. The setoid model thus defined validates function extensionality,
a univalent universe of propositions, and quotient types. Therefore, it provides
a way to bootstrap and ``explain'' extensionality, since the model construction
effectively gives an implementation of various extensionality principles it in
terms of a small, entirely intensional theory.

@@

As any model of Type Theory, the setoid model can be framed categorically as a
category with families (CwF) with extra structure for the various type and term
formers \cite{internal-tt}. The core structure of a CwF can be given as the
following signature:

@@

This definition of types in the setoid model is different from the one in
\cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
difference here is in the use of heterogeneous equality in the definition of
types.

@@

As defined, we can show that the setoid model validates the usual basic type
formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
of strict propositions satisfying propositional extensionality
\cite{setoid99}. Note that we don't need neither identity types nor inductive
types (W-types) in the metatheory to achieve this. However, W-types are needed
when constructing the setoid universe in the next sections.

@@

The setoid model presented in the previous section is peculiar in several
ways. One is that it can be presented in a very small type theoretic
metatheory. The other important aspect is that it is \emph{strict}, namely every
equation of the original syntax holds by definition in the semantics. One
advantage of strict models is that they can be turned into \emph{syntactic
  translations}, in which syntactic objects of the source theory are interpreted
as their counterparts in another \emph{target} theory. In the case of the setoid
model, this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.
\footnote{%
Semantically, this translation corresponds to a model construction, in
particular a functor from the category of models of the target theory to the
category of models of what will be Setoid Type Theory.  Since the setoid
translation is structural in the context component, we can work with models in
the style of categories with families rather than contextual categories.
}

@@

Note that contractibility of singletons holds in SeTT by UIP and the computation
rules of equality on $\Sigma$ types. Hence, Martin-L\"of's J eliminator can also
be derived \todo{add citation}.

@@

As pointed out in the introduction, SeTT is seriously limited by the lack of a
universes internalizing the notion of setoid. Our goal is to extend SeTT with a
universe of setoids; since SeTT is a direct syntactic reflection of the setoid
model, this essentially amounts to showing that a universe of setoids with the
necessary structure and equations can be constructed within the setoid
model. This opens several questions and possible design choices.

A first fundamental consideration has to with the very definition of the setoid
universe: as any type in the setoid model, this universe must be a setoid and
thus come equipped with an equivalence relation. However, unlike the universe of
propositions, a universe of setoids cannot be univalent, since this would force
it to be a groupoid. The obvious choice is therefore to have a non-univalent
universe, and instead define the universe's relation so that it reflects a
simple syntactic equality of codes rather than setoid equivalence.

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section TODO and shown originally in
\cite{setoid99} is that it can be carried out in a very small type theoretic
metatheory, thus providing a way to reduce extensionality to a small intensional
core.
%
We would like to stay faithful to this ideal when constructing this setoid
universe.

A known and established method for defining universes in Type Theory relies on
induction-recursion (IR), a definition schema developed by Dybjer
\cite{ir-dybjer, dybjer-setzer}. Inductive-recursive definitions can be found
throughout the literature, from the already mentioned type theoretic universes,
including the original formulation \`{a} la Tarski by Martin-L\"of
\cite{libretto}, to metamathematical tools like computability predicates.

Although universe constructions in Type Theory---including our own setoid
universe---are naturally presented as inductive-recursive definitions, they
may not necessarily require a metatheory that supports induction-recursion. In
fact, it is possible to reduce some instances of induction-recursion to plain
induction (more specifically, inductive families), including some universe
definitions. We recall this reduction in Section~\ref{ir-universes}.

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such universe eliminator.
%
The question arises because we only define a weak eliminator for our final
inductive-inductive encoding of the setoid universe, for reasons that are
explained in Section~\ref{inductive-universe}. Although not currently needed, a
stronger eliminator might be necessary to justify universe induction. This
problem should not arise in the other encodings of the setoid universe (as given
in Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

Another design choice has to do with how the setoid universe relates to the
other universes. One could have $\Prop$ be a code of the setoid universe, rather
than a type. Moreover, the setoid universes could form a hierarchy, possibly
cumulative.

@@

Induction-recursion is arguably a nicer and more natural way to define
internal universes in Type Theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$).

In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
\emph{carves out} all types in $\Set$ that are in the image of $\El$. In other
words, $\ad{in-U}$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
the type of codes, the definition of $\ad{in-U}$ quite expectedly reflecs the
inductive structure of codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible sets. Hence
this kind of construction requires a metatheory with at least one universe.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes:
%
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
%
in addition to proofs that these are indeed equivalence relations
%
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
%
\begin{align*}
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

@@

This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to understood notions of
induction-recursion like in \cite{dybjer-setzer}. Nevertheless, we would like to
avoid extending the metatheory with any forms of induction-recursion, as per our
goal to keep the metatheory as small and essential as possible.
